Is the direction of greater Granger causal influence the same as the direction of information flow?

### 2015-10-04 paper.tex
Add acknowledgements.

Suppose we try to use the direction of greater Granger causal influence to infer the direction of information flow. As discussed in Section~\ref{sec:intro}, one might expect the direction of information flow -- and hence the direction of greater Granger causality -- to be from the transmitter to the receiver.

@@

In the reverse direction, the Directed Information is $\infty$.
\begin{align*}
	I(0*\widehat{\Theta}^{n-1}\rightarrow X^{n}) &= \sum_{i=0}^{n-1}I(X_{i+1};\widehat{\Theta}^{i}|X^{i}) \\
	I(X_{i+1};\widehat{\Theta}^{i}|X^{i})        &= h(X_{i+1}|X^{i}) - h(X_{i+1}|X^{i},\widehat{\Theta}^{i}) \numberthis \label{eq:dir-info-rev-noiseless}
\end{align*}
The first term in the equation above reduces to
\begin{align*}
	h(X_{i+1}|X^{i}) &= h(\Theta-\widehat{\Theta}_{i}|X^{i}) \\
					 &= h \bigg( \Theta- \bigg( \widehat{\Theta}_{i-1}+\frac{X_{i}+N_{i}}{i} \bigg) \bigg| X^{i} \bigg) \\
					 &\overset{(a)}{=} h \bigg( \Theta- \bigg( (\Theta - X_{i}) + \frac{N_{i}}{i} \bigg) \bigg| X^{i} \bigg) \\
					 &= h \bigg( \frac{N_{i}}{i} \bigg| X^{i} \bigg) \\
					 &= h(N_{i}) -\log(i) \\
					 &= \frac{1}{2}\log(2\pi e\sigma_{N}^{2}) -\log(i)
\end{align*}
where for (a), we have dropped $X_i/i$ from the previous step, since it is conditioned over, and then written $\widehat\Theta_{i-1}$ as $(\Theta - X_i)$. On the other hand, the second term in equation \eqref{eq:dir-info-rev-noiseless} becomes
\begin{align*}
	h(X_{i+1} | X^i, \widehat\Theta^i) &= h(\Theta - \widehat\Theta_i | X^i, \widehat\Theta^i) \\
									   &= h(\Theta | X^i, \widehat\Theta^i) \\
									   &\overset{(a)}{=} h(X_i + \widehat\Theta_{i-1} | X^i, \widehat\Theta^i) \\
									   &= h(0 | X^i, \widehat\Theta^i) \\
									   &\overset{(b)}{=} - \infty
\end{align*}
where for (a) we have expressed $\Theta$ in terms of $\widehat\Theta_{i-1}$ and $X_i$ and for (b), we have used the fact that the differential entropy of a constant (or equivalently, a Gaussian with zero variance) is negative infinity. This means that equation \eqref{eq:dir-info-rev-noiseless} becomes
\begin{align*}
	I(X_{i+1};\widehat{\Theta}^{i}|X^{i}) &= \infty \\
	\Rightarrow I(0*\widehat{\Theta}^{n-1}\rightarrow X^{n}) &= \infty
\end{align*}
as expected.

@@

The Directed Information in the forward direction can be written as
\begin{align*}
	I(X^n \rightarrow \widehat\Theta^n) &= \sum_{i=1}^n{I(\widehat\Theta_i ; X^i | \widehat\Theta^{i-1})} \\
										&= \sum_{i=1}^n{h(\widehat\Theta_i | \widehat\Theta^{i-1}) - h(\widehat\Theta_i | \widehat\Theta^{i-1}, X^i)} \numberthis \label{eq:dir-info-fwd-noisy} \\
										&= \sum_{i=1}^n \bigg( \frac{1}{2} \log(2 \pi e \text{Var}[\Theta - R_{i-1} + N_i | \widehat\Theta^{i-1}]) \\
										& \qquad \qquad - \frac{1}{2}\log(2 \pi e \sigma_N^2) \bigg)
\end{align*}
For a derivation of this, see appendix~\ref{app:dir-info-fwd-noisy} which appears in~\cite{Allerton15Online}. In the reverse direction,
\begin{align*}
	I(0*\widehat\Theta^{n-1} & \rightarrow X^n) = \sum_{i=0}^{n-1}{I(X_{i+1}; \widehat\Theta^i | X^i)} \\
							 &= \sum_{i=0}^{n-1}{h(X_{i+1} | X^i) - h(X_{i+1} | X^i, \widehat\Theta^i)} \numberthis \label{eq:dir-info-rev-noisy} \\
							 &= \frac{1}{2} \log \bigg( 2 \pi e \frac{\sigma_N^2 + \sigma_R^2}{\sigma_R^2} \bigg) \\
							 & + \sum_{i=2}^{n-1} \bigg( \frac{1}{2} \log \bigg( 2 \pi e \text{Var} \bigg[ R_{i-1} - \frac{N_i}{i} - R_i \bigg| X^i \bigg] \bigg) \\
							 & \qquad \qquad - \frac{1}{2} \log( 2 \pi e \sigma_R^2 ) \bigg)
\end{align*}
For a derivation of this, see appendix~\ref{app:dir-info-rev-noisy} which appears in~\cite{Allerton15Online}.

@@

For a given time instant $t$, the $i$\textsuperscript{th} trial is modeled as
\begin{equation*}
	X_t^{(i)} = \sum_{j=1}^p \alpha_j(t) X_{t-j}^{(i)} + \epsilon_t^{(i)}
\end{equation*}
Note that $\alpha_j(t)$ does not depend on $i$. Collecting variables across $N$ trials, we can write the full model in vector form:
\begin{align*}
	\left[ \!\! \begin{array}{c}
			X_t^{(1)} \\
			X_t^{(2)} \\
			\vdots \\
			X_t^{(N)}
	\end{array} \!\! \right]
	&=
	\left[ \! \begin{array}{ccc}
			X_{t-1}^{(1)} & \cdots & X_{t-p}^{(1)} \\
			X_{t-1}^{(2)} & \cdots & X_{t-p}^{(2)} \\
			\vdots & \ddots & \vdots \\
			X_{t-1}^{(N)} & \cdots & X_{t-p}^{(N)} \\
			%X_{t-1}^{(1)} & X_{t-2}^{(1)} & \cdots & X_{t-p}^{(1)} \\
			%X_{t-1}^{(2)} & X_{t-2}^{(2)} & \cdots & X_{t-p}^{(2)} \\
			%\vdots &  & \ddots & \vdots \\
			%X_{t-1}^{(N)} & X_{t-2}^{(N)} & \cdots & X_{t-p}^{(N)} \\
	\end{array} \! \right]
	\left[ \!\! \begin{array}{c}
			\alpha_1(t) \\
			\alpha_2(t) \\
			\vdots \\
			\alpha_p(t)
	\end{array} \!\! \right] +
	%&\qquad \qquad \qquad +
	\left[ \!\! \begin{array}{c}
			\epsilon_t^{(1)} \\
			\epsilon_t^{(2)} \\
			\vdots \\
			\epsilon_t^{(N)}
	\end{array} \!\! \right]
\end{align*}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
	\centering
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point20} \\
	\vspace{5mm}
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point35} \\
	\vspace{5mm}
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point60}
	\caption{Plots for forward and backward directed information computations for ${\sigma_R^2}/{\sigma_N^2}=0.2$ (top), $0.35$ (center) and $0.6$ (bottom). In each plot, curves for directed information in both directions are illustrated for ratios ${\sigma_\theta^2}/{\sigma_N^2} = 0.25, 0.5,$ and $1$. The x-axis is the number of iterations of message-passing between the transmitter and the receiver. For cases when feedback noise variance $\sigma_R^2$ is moderately smaller than feedforward noise variance $\sigma_N^2$, directed information in the reverse link can dominate that in the forward link. With sufficiently many (albeit sometimes large, as illustrated in the top figure) iterations, directed information in forward information starts to dominate. However, the point at which this happens depends on the (often unknown) ratio of noise in these links. }
	\label{fig:numerical}
\end{figure}

@@

As a final remark, we emphasize that this work only demonstrates the error in \emph{interpreting} the direction of causal influence as the direction of \emph{information flow}. We do \emph{not} seek to invalidate much of the neuroscientific work that has been done in this direction; we merely caution against making (what might be construed as hopeful) extrapolations from causal influences as to the manner in which information flows.

@@

The authors would like to thank Nicola Elia for discussions that suggested this possible misunderstanding in use of Granger causality and Directed Information in computational systems. We also thank Rob Kass, Momin Malik and Kun Zhang for useful discussions. Finally, we thank the reviewers of a previous submission of this work to ISIT for their insightful remarks and comments. Praveen Venkatesh is partially supported by a dean's fellowship. %Pulkit Grover is supported by [...].

>>>>

%Suppose we try to use the direction of greater Granger causal influence to infer the direction of information flow. As discussed in Section~\ref{sec:intro}, one might expect the direction of information flow -- and hence the direction of greater Granger causality -- to be from the transmitter to the receiver.

@@

In the reverse direction, the Directed Information is $\infty$.
\begin{align*}
	I(0*\widehat{\Theta}^{n-1}\rightarrow X^{n}) &= \sum_{i=0}^{n-1}I(X_{i+1};\widehat{\Theta}^{i}|X^{i}) \\
	I(X_{i+1};\widehat{\Theta}^{i}|X^{i})        &= h(X_{i+1}|X^{i}) - h(X_{i+1}|X^{i},\widehat{\Theta}^{i}) \numberthis \label{eq:dir-info-rev-noiseless}
\end{align*}
The first term in the equation above reduces to
\begin{align*}
	h(X_{i+1}|X^{i}) &= h(\Theta-\widehat{\Theta}_{i}|X^{i}) \\
					 &= h \bigg( \Theta- \bigg( \widehat{\Theta}_{i-1}+\frac{X_{i}+N_{i}}{i} \bigg) \bigg| X^{i} \bigg) \\
					 &\overset{(a)}{=} h \bigg( \Theta- \bigg( (\Theta - X_{i}) + \frac{N_{i}}{i} \bigg) \bigg| X^{i} \bigg) \\
					 &= h \bigg( \frac{N_{i}}{i} \bigg| X^{i} \bigg) \\
					 &= h(N_{i}) -\log(i) \\
					 &= \frac{1}{2}\log(2\pi e\sigma_{N}^{2}) -\log(i)
\end{align*}
where for (a), we have dropped $X_i/i$ from the previous step, since it is conditioned over, and then written $\widehat\Theta_{i-1}$ as $(\Theta - X_i)$. On the other hand, the second term in equation \eqref{eq:dir-info-rev-noiseless} becomes
\begin{align*}
	h(X_{i+1} | X^i, \widehat\Theta^i) &= h(\Theta - \widehat\Theta_i | X^i, \widehat\Theta^i) \\
									   &= h(\Theta | X^i, \widehat\Theta^i) \\
									   &\overset{(a)}{=} h(X_i + \widehat\Theta_{i-1} | X^i, \widehat\Theta^i) \\
									   &= h(0 | X^i, \widehat\Theta^i) \\
									   &\overset{(b)}{=} - \infty
\end{align*}
where for (a) we have expressed $\Theta$ in terms of $\widehat\Theta_{i-1}$ and $X_i$ and for (b), we have used the fact that the differential entropy of a constant (or equivalently, a Gaussian with zero variance) is negative infinity. This means that equation \eqref{eq:dir-info-rev-noiseless} becomes
\begin{align*}
	I(X_{i+1};\widehat{\Theta}^{i}|X^{i}) &= \infty \\
	\Rightarrow I(0*\widehat{\Theta}^{n-1}\rightarrow X^{n}) &= \infty
\end{align*}
%as expected.

@@

The Directed Information in the forward direction can be written as
\begin{align*}
	I(X^n \rightarrow \widehat\Theta^n) &= \sum_{i=1}^n{I(\widehat\Theta_i ; X^i | \widehat\Theta^{i-1})} \\
										&= \sum_{i=1}^n{h(\widehat\Theta_i | \widehat\Theta^{i-1}) - h(\widehat\Theta_i | \widehat\Theta^{i-1}, X^i)} \numberthis \label{eq:dir-info-fwd-noisy} \\
										&= \sum_{i=1}^n \bigg( \frac{1}{2} \log(2 \pi e \text{Var}[\Theta - R_{i-1} + N_i | \widehat\Theta^{i-1}]) \\
										& \qquad \qquad - \frac{1}{2}\log(2 \pi e \sigma_N^2) \bigg)
\end{align*}
For a derivation of this, see appendix~\ref{app:dir-info-fwd-noisy} which appears in~\cite{Allerton15Online}.

In the reverse direction,
\begin{align*}
	I(0*\widehat\Theta^{n-1} & \rightarrow X^n) = \sum_{i=0}^{n-1}{I(X_{i+1}; \widehat\Theta^i | X^i)} \\
							 &= \sum_{i=0}^{n-1}{h(X_{i+1} | X^i) - h(X_{i+1} | X^i, \widehat\Theta^i)} \numberthis \label{eq:dir-info-rev-noisy} \\
							 &= \frac{1}{2} \log \bigg( 2 \pi e \frac{\sigma_N^2 + \sigma_R^2}{\sigma_R^2} \bigg) \\
							 & + \sum_{i=2}^{n-1} \bigg( \frac{1}{2} \log \bigg( 2 \pi e \text{Var} \bigg[ R_{i-1} - \frac{N_i}{i} - R_i \bigg| X^i \bigg] \bigg) \\
							 & \qquad \qquad - \frac{1}{2} \log( 2 \pi e \sigma_R^2 ) \bigg)
\end{align*}
For a derivation of this, see appendix~\ref{app:dir-info-rev-noisy} which appears in~\cite{Allerton15Online}.

@@

For a given time instant $t$, the $i$\textsuperscript{th} trial is modeled as
\begin{equation*}
	X_t^{(i)} = \sum_{j=1}^p \alpha_j(t) X_{t-j}^{(i)} + \epsilon_t^{(i)}
\end{equation*}
Note that $\alpha_j(t)$ does not depend on $i$. Collecting variables across $N$ trials, we can write the full model in vector form:

\begin{align*}
	\left[ \!\! \begin{array}{c}
			X_t^{(1)} \\
			X_t^{(2)} \\
			\vdots \\
			X_t^{(N)}
	\end{array} \!\! \right]
	&=
	\left[ \! \begin{array}{ccc}
			X_{t-1}^{(1)} & \cdots & X_{t-p}^{(1)} \\
			X_{t-1}^{(2)} & \cdots & X_{t-p}^{(2)} \\
			\vdots & \ddots & \vdots \\
			X_{t-1}^{(N)} & \cdots & X_{t-p}^{(N)} \\
			%X_{t-1}^{(1)} & X_{t-2}^{(1)} & \cdots & X_{t-p}^{(1)} \\
			%X_{t-1}^{(2)} & X_{t-2}^{(2)} & \cdots & X_{t-p}^{(2)} \\
			%\vdots &  & \ddots & \vdots \\
			%X_{t-1}^{(N)} & X_{t-2}^{(N)} & \cdots & X_{t-p}^{(N)} \\
	\end{array} \! \right]
	\left[ \!\! \begin{array}{c}
			\alpha_1(t) \\
			\alpha_2(t) \\
			\vdots \\
			\alpha_p(t)
	\end{array} \!\! \right] +
	%&\qquad \qquad \qquad +
	\left[ \!\! \begin{array}{c}
			\epsilon_t^{(1)} \\
			\epsilon_t^{(2)} \\
			\vdots \\
			\epsilon_t^{(N)}
	\end{array} \!\! \right]
\end{align*}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
	\centering
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point20} \\
	%\vspace{5mm}
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point35} \\
	%\vspace{5mm}
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point60}
	\caption{Plots for forward and backward directed information computations for ${\sigma_R^2}/{\sigma_N^2}=0.2$ (top), $0.35$ (center) and $0.6$ (bottom). In each plot, curves for directed information in both directions are illustrated for ratios ${\sigma_\theta^2}/{\sigma_N^2} = 0.25, 0.5,$ and $1$. The x-axis is the number of iterations of message-passing between the transmitter and the receiver. For cases when feedback noise variance $\sigma_R^2$ is moderately smaller than feedforward noise variance $\sigma_N^2$, directed information in the reverse link can dominate that in the forward link. With sufficiently many (albeit sometimes large, as illustrated in the top figure) iterations, directed information in forward information starts to dominate. However, the point at which this happens depends on the (often unknown) ratio of noise in these links. }
	\label{fig:numerical}
\end{figure}

@@

As a final remark, we emphasize that this work only demonstrates the error in \emph{interpreting} the direction of causal influence as the direction of \emph{information flow}. We do \emph{not} seek to invalidate much of the neuroscientific work that has been done in this direction; we merely caution against making (what might be construed as hopeful) extrapolations from causal influences to information flows.

@@

The authors would like to thank Nicola Elia for discussions that suggested this possible misunderstanding in use of Granger causality and Directed Information in computational systems. We also thank Rob Kass, Bruno Sinopoli, Momin Malik and Kun Zhang for useful discussions. Finally, we thank the reviewers of a previous conference submission of this work for their insightful remarks and comments. We also thank the CMU BrainHUB initiative for support through a Proseed-BrainHUB seed grant. Pulkit Grover was supported through a CMU startup grant, and Praveen Venkatesh through a CMU CIT Dean's fellowship and a CMU Presidential Fellowship.
